---
title: "CDC Data Exercise"
author: "Emma Hardin-Parker"
date: "02/07/2024"
---

### Data Introduction and Loading

I accessed the data for this assignment at this [link](https://data.cdc.gov/NCHS/NCHS-VSRR-Quarterly-provisional-estimates-for-sele/76vv-a7x8/about_data). I chose a data set from the National Center for Health Statistics (NCHS) that provides data on quarterly provisional estimates for selected birth indicators. These indicators include teen birth rate, low birth weight, NICU admissions, pre-term births, etc. These indicators are stratified by race/ethnicity and age groups. There is an overall Rate column that is subsetted into "per 100 births" or "per 1,000 population."

I am now going to load any required packages along with the original, uncleaned data set.

```{r}
library(readr)
library(here)
library(tidyverse)
library(ggplot2)

nchs_data <- read.csv("~/Desktop/NCHS_-_VSRR_Quarterly_provisional_estimates_for_selected_birth_indicators_20240206.csv")
View(nchs_data)
```

### Initial Data Exploration and Cleaning

I am now going to get a quick look at the data by using the str(), summary(), and glimpse() functions.

```{r}
str(nchs_data)
summary(nchs_data)
glimpse(nchs_data)
```

The data set has 1,110 observations and 8 variables. At first glance there doesn't seem to be any NA values, but I am going to check using the NA values. After print nchs_na, I found that there were no NA values, however, I took that command out as it printed "FALSE" for every observation in the data set. Also, I noticed in the Significant column, that significant values are coded with asteriks. I could imagine that would make future exploratory analyses difficult, so I plan to recode significant values to 1 and non-significant values to 0.

```{r}
nchs_na <- is.na(nchs_data)

nchs_data <- nchs_data %>%
  mutate(Significant = as.character(Significant))

nchs_recode <- nchs_data %>%
  mutate(Significance_Status = case_when(
    Significant == "*" ~ 1,
    Significant == "" ~ 0,
    TRUE ~ NA_integer_  
  ))
```

To accomplish this I used the is.na() function to determine if there were any NA values in the data set. I then mutated the Significant column to treat the observations as characters (\* or missing) by using as.character(). I then mutated a new variable, Significance_Status to recode the missing values as zeroes and the asteriks as ones.

### Exploratory Data Analysis

I am now going to create an exploratory plot to see which race/ethnicity category had the largest birth rate by Year & Quarter.

```{r}
filtered_race <- nchs_recode %>%
  filter(Race.Ethnicity.Category != "All races and origins")  # Remove rows where 'Race' is "All"


birth_rate_bar <- filtered_race %>%
  group_by(Year.and.Quarter) %>%
  filter(Topic == "Birth Rates") %>% 
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate, fill = Race.Ethnicity.Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Birth Rate Rate by Age Category and Race",
       x = "Age Category / Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  theme(axis.text.x = element_text(angle = 60, hjust = 1))  

hispanic_rate_bar <- nchs_recode %>%
  filter(Race.Ethnicity.Category == "Hispanic") %>%
  filter(Topic == "Birth Rates") %>%
  group_by(Year.and.Quarter) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate, fill = Race.Ethnicity.Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Birth Rate Rate by Age Category and Race",
       x = "Age Category / Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) 

white_rate_bar <- nchs_recode %>%
  filter(Race.Ethnicity.Category == "Non-Hispanic White") %>%
  filter(Topic == "Birth Rates") %>%
  group_by(Year.and.Quarter) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Birth Rate Rate by Age Category and Race",
       x = "Age Category / Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_color_viridis_d()

black_rate_bar <- nchs_recode %>%
  filter(Race.Ethnicity.Category == "Non-Hispanic Black") %>%
  filter(Topic == "Birth Rates") %>%
  group_by(Year.and.Quarter) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Birth Rate Rate by Age Category and Race",
       x = "Age Category / Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  scale_color_viridis_b() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 


print(hispanic_rate_bar)
print(black_rate_bar)
print(white_rate_bar)
```

I am now going to create plots exploring the rate of low birthweight (per 100 births) by race

```{r}
low_birth <- filtered_race %>%
  filter(Topic.Subgroup == "Low Birthweight Rates") %>%
  group_by(Race.Ethnicity.Category) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate, fill = Race.Ethnicity.Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Low Birthweight Rates by Race",
       x = "Age Category / Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  scale_color_identity() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(low_birth)
```

I am now going to create a similar plot concerning preterm birth rates per 100 births by race.

```{r}
pre_birth <- filtered_race %>%
  filter(Topic.Subgroup == "Preterm Birth Rates") %>%
  group_by(Race.Ethnicity.Category) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate, fill = Race.Ethnicity.Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Preterm Birth Rates by Race",
       x = "Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  scale_color_identity() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

print(pre_birth)
```

I am now going to create a similar plot concerning NICU admission rates per 100 births by race.

```{r}
NICU <- filtered_race %>%
  filter(Topic.Subgroup == "NICU Admission Rates") %>%
  group_by(Race.Ethnicity.Category) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate, fill = Race.Ethnicity.Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "NICU Admission Rates by Age Category and Race",
       x = "Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  scale_color_identity() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

print(NICU)
```

I am now going to create a similar plot exploring Teen Birth Rates by Age Category and Race

```{r}
teen_birth <- filtered_race %>%
  filter(Topic.Subgroup == "Teen Birth Rates") %>%
  group_by(Race.Ethnicity.Category) %>%
  ggplot(aes(x = interaction(Indicator, Race.Ethnicity.Category), y = Rate, fill = Race.Ethnicity.Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = " Teen Birth Rate Rate by Age Category and Race",
       x = "Age Category / Race",
       y = "Rate") +
  scale_x_discrete(labels = function(x) gsub("\\.", "-", x)) +  
  scale_color_viridis_b() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

print(teen_birth)
```
